Solar wind, F10.7, and geomagnetic activity relationship to the equatorial plasma mass density at geosynchronous orbit

### 2016-10-18 paper/paper.production.tex
Paper diff in new file, figures split into pdfs

The dependence of the plasmapause location on geomagnetic activity was observed as early as \cite{Carpenter1966} and there are well--validated models of the plasmapause location as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}; \cite{LemaireEarthsPlasmasphere}) with similar parameterization.

The dependence of the plasma density outside of the nominal plasmapause location on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of ion density found a significant dependence of the plasmapause location on geomagnetic activity (\cite{Chappell1970}; \cite{Maynard1977}; \cite{Carpenter1992}), but the relationship between geomagnetic activity and density outside of the plasmapause was not explicitly considered.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 when CRRES was in an elliptical, near equatorial orbit and primarily beyond the nominal plasmapause. They found that the local average ion mass, $M$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and larger 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$ had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of [12:00, 18:00] with most observations at $L$ between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and $K_p$ and the field line distribution of plasma mass density, $\rho$, using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a field line near the equator was found for lower $D_{st}$ and larger $K_p$, but the equatorial value of $\rho$ was found to be similar for $D_{st}$ bins of [-142, -31] and [-31, 37] nT and $K_p$ bins of [1.5,3.4] and [3.4, 5.9].

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density was estimated using the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a storm-time magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)^{1/2}$.  They found a high correlation ($\sim 0.94$) between 27-day averages of $F10.7$ and 27-day medians of $\rho_{eq}$.

\cite{Takahashi2010} noted that downward drops in the $D_{st}$ index coincided with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ increases after $D_{st}$ minimum, two had $\rho_{eq}$ spikes before the minimum, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, the epoch average of $D_{st}$ for storms with a minimum $D_{st} < -50$~nT showed an enhancement in $\rho_{eq}$ the same day as minimum $D_{st}$ in the epoch averages.

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation at geosynchronous distances was not calculated, the data presented for four events show that the enhancements in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The above results indicate that (1) the mass density at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind variables that may drive or be associated with the changes.

%There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, and geomagnetic and solar wind parameters also correlate with F10.7, (3) the dependence of mass density on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF, corresponding to low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of mass density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in mass density associated with strong southward IMF and enhanced magnetospheric activity (\cite{Takahashi2010}; \cite{Yao2008}).

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{figures/ccplot-GOES6.eps}
\caption{Top: 27-day non-overlapping medians of $F_{10.7}$ and $\log_{10}(\rho_{eq})$ from GOES 6. Bottom: Scatter plot of $\log_{10}(\rho_{eq})$ and $F_{10.7}$ using medians in non-overlapping hour, day, and 27-day windows.}
\label{fig:ccplot}
\end{figure}
\clearpage

\begin{figure}[h]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dst-50-tak-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-day-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using daily medians. Top: Events in the interval 1989-1991; compare to \cite{Takahashi2010} Figure~11. Bottom: Events in the interval 1983-1991.}
\label{fig:DailyAveragedDstEvents}
\end{figure}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{figures/stormavs-dst-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events with the constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.  Bottom: Same as Top except without constraint.}
\label{fig:HourlyAveragedDstEvents}
\end{figure}

\clearpage
\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case1-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case1-t025-tf30-GOES6.eps}
  \caption{$D_{st}$ events of the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $D_{st}$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:BzBinnedDstEvents}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case1-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case1-t025-tf30-GOES6.eps}
  \caption{$D_{st}$ events of the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $F10.7$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:F107BinnedDstEvents}
\end{figure}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-mass-gt20-GOES6.eps}
\caption{$\rho_{eq} >$ 20~amu/cm$^3$ events from GOES 6 using hourly medians.}
\label{fig:HourlyAveragedRhoEvents}
\end{figure}

\clearpage
\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case24-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedB_z-case24-t025-tf30-GOES6.eps}
  \caption{$\rho_{eq}$ events of Figure~\ref{fig:HourlyAveragedRhoEvents} separated by the average $B_z$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:BzBinned}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case24-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedF_{10.7}-case24-t025-tf30-GOES6.eps}
  \caption{$\rho_{eq}$ events of Figure~\ref{fig:HourlyAveragedRhoEvents} separated by the average $F10.7$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:F107Binned}
\end{figure}

>>>>

The dependence of the plasmapause location on geomagnetic activity was observed as early as \citet{Carpenter1966} and there are well--validated models of the plasmapause location as a function of $L$, $MLT$, and geomagnetic activity (\citet{LemaireEarthsPlasmasphere}, \citet{Moldwin2002ModelPlasmapause}, \citet{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\citet{Gallagher1988EmpiricalModelPlasmasphere}; \citet{LemaireEarthsPlasmasphere}) with similar parameterization.

The dependence of the plasma density outside of the nominal plasmapause location on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\citet{Takahashi2006}; \citet{Takahashi2010}; \citet{Denton2016}). Earlier works that considered profiles of ion density found a significant dependence of the plasmapause location on geomagnetic activity (\citet{Chappell1970}; \citet{Maynard1977}; \citet{Carpenter1992}), but the relationship between geomagnetic activity and density outside of the plasmapause was not explicitly considered.

\citet{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 when CRRES was in an elliptical, near equatorial orbit and primarily beyond the nominal plasmapause. They found that the local average ion mass, $M$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and larger 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$ had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of [12:00, 18:00] with most observations at $L$ between 5 and 7 $R_E$.

\citet{Denton2006} found a weak relationship between $D_{st}$ and $K_p$ and the field line distribution of plasma mass density, $\rho$, using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a field line near the equator was found for lower $D_{st}$ and larger $K_p$, but the equatorial value of $\rho$ was found to be similar for $D_{st}$ bins of [-142, -31] and [-31, 37] nT and $K_p$ bins of [1.5,3.4] and [3.4, 5.9].

\citet{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density was estimated using the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a storm-time magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)^{1/2}$.  They found a high correlation ($\sim 0.94$) between 27-day averages of $F10.7$ and 27-day medians of $\rho_{eq}$.

\citet{Takahashi2010} noted that downward drops in the $D_{st}$ index coincided with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ increases after $D_{st}$ minimum, two had $\rho_{eq}$ spikes before the minimum, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, the epoch average of $D_{st}$ for storms with a minimum $D_{st} < -50$~nT showed an enhancement in $\rho_{eq}$ the same day as minimum $D_{st}$ in the epoch averages.

\citet{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation at geosynchronous distances was not calculated, the data presented for four events show that the enhancements in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The above results indicate that (1) the mass density at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \citet{Takahashi2010} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind variables that may drive or be associated with the changes.

%There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, and geomagnetic and solar wind parameters also correlate with F10.7, (3) the dependence of mass density on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF, corresponding to low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of mass density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in mass density associated with strong southward IMF and enhanced magnetospheric activity (\citet{Takahashi2010}; \citet{Yao2008}).

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p02a.eps}
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p02b.eps}
\caption{Top: 27-day non-overlapping medians of $F_{10.7}$ and $\log_{10}(\rho_{eq})$ from GOES 6. Bottom: Scatter plot of $\log_{10}(\rho_{eq})$ and $F_{10.7}$ using medians in non-overlapping hour, day, and 27-day windows.}
\label{fig:ccplot}
\end{figure}
\clearpage

\begin{figure}[h]
\centering
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p03a.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p03b.eps}
\caption{$D_{st}$ events from GOES 6 using daily medians. Top: Events in the interval 1989-1991; compare to \citet{Takahashi2010} Figure~11. Bottom: Events in the interval 1983-1991.}
\label{fig:DailyAveragedDstEvents}
\end{figure}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p04a.eps}
\\
\rule[1ex]{5cm}{1pt}
\\
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p04b.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events with the constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.  Bottom: Same as Top except without constraint.}
\label{fig:HourlyAveragedDstEvents}
\end{figure}

\clearpage
\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p05a.eps}
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p05b.eps}
  \caption{$D_{st}$ events of the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $D_{st}$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:BzBinnedDstEvents}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p06a.eps}
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p06b.eps}
  \caption{$D_{st}$ events of the bottom panel of Figure~\ref{fig:HourlyAveragedDstEvents} separated by the average $F10.7$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:F107BinnedDstEvents}
\end{figure}

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p07.eps}
\caption{$\rho_{eq} >$ 20~amu/cm$^3$ events from GOES 6 using hourly medians.}
\label{fig:HourlyAveragedRhoEvents}
\end{figure}

\clearpage
\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p08a.eps}
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p08b.eps}
  \caption{$\rho_{eq}$ events of Figure~\ref{fig:HourlyAveragedRhoEvents} separated by the average $B_z$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:BzBinned}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p09a.eps}
  \includegraphics[scale=0.40]{UsedFigures/2016SW001507R-p09b.eps}
  \caption{$\rho_{eq}$ events of Figure~\ref{fig:HourlyAveragedRhoEvents} separated by the average $F10.7$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:F107Binned}
\end{figure}
